---
title: "Covariate-adjusted statistical dependence representation through partial copulas: bounds and new insights"
author:
- Vinícius Litvinoff Justus (Carlos III University of Madrid, Spain)
- Felipe Fontana Vieira (Ghent University, Belgium)
date: "March 2026"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    citation_package: natbib
    toc: false
  html_document:
    toc: false
    df_print: paged
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsthm}
- \newtheorem{definition}{Definition}
- \newtheorem{proposition}{Proposition}
- \newtheorem{corollary}{Corollary}
- \newtheorem{remark}{Remark}
- \newtheorem{example}{Example}
- \newcommand{\UX}{U_X}
- \newcommand{\UY}{U_Y}
- \newcommand{\independent}{\perp\!\!\!\perp}
link-citations: true
abstract: "In this paper, we revisit the notion of partial copula, originally introduced to test conditional independence, highlighting its capability to represent the dependence between two random variables after removing their dependence with a covariate. Building upon results previously presented in the literature, we show that partial copulas can be seen as a nonlinear analogue of partial correlation. Then, we prove several results showing how dependence properties of the conditional copulas constrain the form of the partial copula. Finally, a simulation study is conducted to illustrate the results and to show the potential of partial copula as a way to describe covariate-adjusted statistical dependence. This highlights the potential of the method to be used in causal inference problems and recover the true sign of a causal effect.\n"
keywords: Copula theory; Rosenblatt transformation; partial correlation; causal inference
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = 'center')

library(rvinecopulib)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

set.seed(123)
n <- 5000
```

# Introduction

Quantifying the strength of the statistical association between two random variables, controlled by one or more variables (usually called covariates or control variables), is an old problem in statistics and widely discussed in the context of linear models and regression coefficients [@kutner2005; @baba2004; @kim2015]. It is well established in the statistical literature that, under wide conditions, conditionally independent random variables are unconditionally correlated\footnote{In fact, latent variables ensuring conditional independence are a common tool in statistics to ``create'' or ``explain'' dependence between variables - see, for instance, factor models \citep{joe2014, Bork2017, Nguyen2020} and de Finetti's representation theorem \citep{oneill2009}.} [see theorem 2 in @oneill2009] - therefore, in the context of causal models, two variables causally unrelated may be correlated if they are both affected by a third variable, called confounder [@pearl2009; @pearl2016]. For this reason, the analysis of conditional and partial associations is of great importance in statistical analysis of observational data; in the context of regression analysis, this is done adding control variables [@kutner2005], and, in the context of correlation analysis, this is done through partial correlation [@baba2004; @kim2015]. A brief introduction on how to choose control variables properly (from a causal viewpoint) is presented in @cinelli2022.

In the same way that using a single numerical index (such as the expectation) is not as informative as knowing the distribution of a variable, it is also true that correlation coefficients fail in capturing several aspects of the dependence structure of a random vector - for this reason, copula theory offers a much richer way to study statistical dependence^[@embrechts2002 presents an example of two bivariate distributions share same marginal distributions and same Pearson's correlation value, but the tail dependence is not the same because the copulas are not the same. This exemplifies the limitation of using Pearson's correlation coefficient to summarise the dependence structure.]. Copulas are functions such that, given the copula and the univariate (marginal) distributions of a random vector, the joint distribution of the variables is fully characterized - therefore, they fully characterize the dependence structure behind a random vector [@nelsen2006; @joe2014]. In fact, it has been proved that copulas offer sufficient and necessary conditions for independence [Theorem 2.4.2 in @nelsen2006] and conditional independence [Theorem 1 in @gonzalezlopez2024]. Therefore, having the notion of partial copula is of great relevance to fully study the stochastic dependence between two variables after controlling for some confounder, because they are a natural extension of the notion of partial correlation.

Partial copulas have been introduced in @bergsma2004 as a tool to construct a conditional independence test [which also have been explored in @bergsma2011; @patra2016; and @petersen2021], and their use in vine copula and stochastic process modeling have been discussed in the literature [@bladt2022; @bauer2016; @spanhel2019]. However, to the best of our knowledge, the use of this notion as a way to formally study covariate-adjusted dependence have not been explored enough in the literature [despite the existence of some references exploring related matters, see @gijbels2015 and @spanhel2016]. In the context of linear models, the use of copulas to enable causal inference was proposed in @park2012 and further studied in @falkenstrom2023; the method proposed by the authors is a data-driven method that attempts to remove the confounder effect without measuring the confounding variable directly (which may require strong untestable assumptions). More recently, the use of vine copulas in causal inference was considered by @justus2024 and @czado2025 - however, to the best of our knowledge, the use of partial copulas to quantify causal effects is not yet properly explored in the literature.

Our main contributions are twofold. First, we give a practical interpretation of the notion of partial copula as a nonlinear analogue of partial correlation, suggesting that its use should not be limited to conditional independence testing. Second, we prove several results showing how properties of the conditional copulas - such as Kolmogorov distance dependence, quadrant positive dependence and Spearman's correlation - constrain the form of the partial copula. Additionally, the results are illustrated through simulations.

The paper is organized as follows. Next section provides all the necessary background to the study of partial copulas: Rosenblatt transformations, probability integral transform and copula functions are introduced. Section 3 is divided in three subsections, the first two discussing the interpretation of partial copula and the last one introducing new results connecting conditional and partial copulas. A simulation study is conducted in Section 4, and final considerations are given in Section 5.

# Background

## Copulas

Copulas are multivariate functions $[0,1]^d \to [0,1]$ used to describe the dependence structure of random vectors of dimension $d \in \mathbb{N}$. The definition in the bivariate case is presented below.

\begin{definition}\label{def:2copula}
A 2-copula is a function $C: [0,1]^2 \to [0,1]$ with the following properties:

(i) $C(u,0) = C(0,v) = 0$, and $C(u,1) = u$, $C(1,v) = v$, $\forall u,v: u,v \in [0,1]$;

(ii) for every $u_i, v_i \in [0,1]$, $i = 1,2$, such that $u_1 \leq u_2$ and $v_1 \leq v_2$,
$$C(u_2, v_2) - C(u_2, v_1) - C(u_1, v_2) + C(u_1, v_1) \geq 0.$$
\end{definition}

Note that Definition \ref{def:2copula} is equivalent to the definition of 2-variate cumulative distribution function (CDF) whose marginal distributions are uniform in $[0,1]$ range. Therefore, given two any random variables $U, V$ such that $U \sim U(0,1)$ and $V \sim U(0,1)$, the cumulative distribution function $P(U \leq u, V \leq v)$ is necessarily a copula (because it satisfies Definition \ref{def:2copula}).

Sklar's theorem [@sklar1959; @nelsen2006] guarantees that, given two any random variables $X$ and $Y$, it holds that
$$P(X \leq x, Y \leq y) = C(F_X(x), F_Y(y)),$$
where $F_X$ and $F_Y$ are, respectively, the cumulative distribution functions of $X$ and $Y$, and $C$ is a copula - if the variables are continuous, then $C$ is unique and it may be called "the copula of $(X,Y)$". It's possible to note that the copula of $(X,Y)$ is the CDF of $(F_X(X), F_Y(Y))$.

The representation provided by Sklar's theorem shows that copulas are a "marginal-free" way to study the dependence between two (or more) random variables, which makes clear that this is a central notion in the study of stochastic dependence. For further arguments regarding the importance of copula theory in the study of stochastic dependence, see @embrechts2002 and see also the monographs from @nelsen2006 and @joe2014.

## Probability integral transform and Rosenblatt transforms

An important result in both copula theory and computational statistics is the so-called "probability integral transform" [@int-transf], presented below. This result ensures that a cumulative distribution function makes a variable uniformly distributed when it is used as a transformation applied in the random variable. Given that copulas are cumulative distribution functions of variables uniformly distributed in $[0,1]$, the importance of this result for copula theory seems evident.

\begin{proposition}[Probability Integral Transform]
Let $F$ be an absolutely continuous cumulative distribution function, and let $F^{-1}$ be the inverse function of $F$. Then,
\begin{itemize}
\item If $U \sim U(0,1)$, then $F^{-1}(U)$ is a random variable with cumulative distribution function $F$ (which will be denoted as $F^{-1}(U) \sim F$).
\item Conversely, if $X \sim F$, then $F(X)$ is a random variable with uniform distribution on $(0,1)$, this is, $F(X) \sim U(0,1)$.
\end{itemize}
\end{proposition}

An even more general result is available in @rosenblatt1952: given a sequence of random variables $X_1, \ldots, X_n$, the random variables
$$F_1(X_1), \, F_{X_2|X_1}(X_2|X_1), \, \ldots, \, F_{X_n|X_1,\ldots,X_{n-1}}(X_n|X_1, \ldots, X_{n-1})$$
are independent and uniformly distributed in the $[0,1]$ range. These functions are usually called "Rosenblatt transforms" and they can be seen as a kind of generalization of the probability integral transform to the multivariate case. In statistical literature, those transformations already have been employed to test conditional independence [@bergsma2004; @Song; @patra2016] and, more recently, to construct a nonparametric notion of error term in regression models [@patra2016]. A central result connecting Rosenblatt transforms and conditional independence is given by next proposition.

\begin{proposition}\citep{patra2016}
Let $(X, Y, Z)$ be an absolutely continuous random vector. Therefore, $X \independent Y | Z$ if and only if $F_{X|Z}(X|Z)$, $F_{Y|Z}(Y|Z)$ and $Z$ are mutually independent.
\end{proposition}

This result suggests that studying the joint distribution of $F_{X|Z}(X|Z)$ and $F_{Y|Z}(Y|Z)$ is a natural way to study the dependence between $X$ and $Y$ adjusted by $Z$, because those two quantities would be independent in the particular case that they are conditionally independent given $Z$ (therefore, this joint distribution would remove the "spurious association" induced by $Z$). In the context of causal inference, if $Z$ is the confounder of $(X,Y)$, the two variables would be conditionally independent given $Z$, provided they are causally unrelated (see considerations about fork models in [@pearl2009; @pearl2016].

# The partial copula

## Definition and interpretation

This section presents the notion of partial copula, originally introduced in @bergsma2004.

\begin{definition}
Let $(X, Y, Z)$ be an absolutely continuous random vector. The partial copula of $(X, Y)$ controlled by $Z$, which we denote by $C_{X,Y;Z}$, is defined as
$$C_{X,Y;Z}(x, y) = P(\UX \leq x, \UY \leq y),$$
where $\UX = F_{X|Z}(X|Z)$ and $\UY = F_{Y|Z}(Y|Z)$.
\end{definition}

\begin{remark}\label{rem:remark-2}
Note that $C_{X,Y;Z}$ fully satisfies the definition of copula (Definition \ref{def:2copula}), because $C_{X,Y;Z}$ is the cumulative distribution function of two random variables uniformly distributed in $[0,1]$.
\end{remark}

The partial copula $C_{X,Y;Z}$ may be seen - in a similar way to partial correlation - as a notion that represents the stochastic dependence between $(X, Y)$ after removing the effect of a covariate $Z$. Further, differently from the notion of conditional copula, $C_{X,Y;Z}$ does not depend on the specification of a specific value $z$ of $Z$.

Next result shows that the partial copula $C_{X,Y;Z}$ can be written as a mixture of all conditional copulas $C_{X,Y|Z=z}$; this result, in addition to providing an analytic representation for the partial copula, highlights the capability of the concept in capturing conditional dependence.

\begin{proposition}\label{prop:partial-copula}
Let $(X, Y, Z)$ be a continuous random vector. The partial copula $C_{X,Y;Z}$ can be written as
$$C_{X,Y;Z}(x, y) = \int_{-\infty}^{\infty} C_{X,Y|Z=z}(x, y) f_Z(z) dz,$$
where $C_{X,Y|Z=z}$ is the conditional copula of $(X, Y)$ given $Z = z$.
\end{proposition}

\begin{proof}
Note that
\begin{align*}
C_{X,Y;Z}(x, y) &= P(\UX \leq x, \UY \leq y) \\
&= \int_{-\infty}^{\infty} P(\UX \leq x, \UY \leq y | Z = z) f_Z(z) dz \\
&= \int_{-\infty}^{\infty} P(F_{X|Z}(X|Z) \leq x, F_{Y|Z}(Y|Z) \leq y | Z = z) f_Z(z) dz \\
&= \int_{-\infty}^{\infty} P(F_{X|Z}(X|z) \leq x, F_{Y|Z}(Y|z) \leq y | Z = z) f_Z(z) dz \\
&= \int_{-\infty}^{\infty} C_{X,Y|Z=z}(x, y) f_Z(z) dz,
\end{align*}
where the third equality comes from applying substitution principle for conditional distributions \citep[Proposition 4.1 of][]{james1981}; and last equality comes from applying Sklar's theorem \citep{sklar1959} in $P(F_{X|Z}(X|z) \leq x, F_{Y|Z}(Y|z) \leq y | Z = z)$.

This application of Sklar's theorem is justified because $P(F_{X|Z}(X|z) \leq x, F_{Y|Z}(Y|z) \leq y | Z = z)$ is a cumulative distribution function for any $z \in \mathbb{R}$; Furthermore, $F_{X|Z}(X|z)$ and $F_{Y|Z}(Y|z)$ are, respectively, strictly increasing transformations of $x$ and $y$, because they are cumulative distribution functions for any $z \in \mathbb{R}$; then, from Theorem 2.4.3 in \citep{nelsen2006}, the conditional copula identified by Sklar's theorem in $P(F_{X|Z}(X|z) \leq x, F_{Y|Z}(Y|z) \leq y | Z = z)$ is the same as the conditional copula of $P(X \leq x, Y \leq y | Z = z)$. Therefore, the last equality is proved taking into account that the univariate conditional distributions are uniform in $[0,1]$.
\end{proof}

Proof of Proposition 3 ensures a formal justification to a similar expression that appears in proof of Proposition 1 in @gijbels2015.

\begin{remark}
A straightforward adaptation of the proof of Proposition 3 \citep[see also Equation 2.1 in][]{joe1996families} shows that, given a continuous random vector $(X,Y,Z)$,
\begin{align*}
P(X \leq x, Y \leq y) &= \int_{-\infty}^{\infty} P(X \leq x, Y \leq y | Z = z) f_Z(z) dz \\
&= \int_{-\infty}^{\infty} C_{X,Y|Z=z}(F_{X|Z}(x|z), F_{Y|Z}(y|z)) f_Z(z) dz.
\end{align*}
Therefore, the partial copula can also be seen as the distribution of $(X,Y)$ under the assumption that each variable is independent of $Z$ but keeping the same conditional copulas as in the true distribution of $(X,Y,Z)$.
\end{remark}

An interesting consequence of the last proposition - summarized in the following corollary - is that, when the "simplifying assumption" [commonly used in Vine copula modeling, see @czado2022] is adopted, then the partial copula coincides with the conditional copula.

\begin{corollary}\label{cor:partial-conditional-copula}
Let $(X, Y, Z)$ be a continuous random vector. If the conditional copula of $(X, Y)$ given $Z = z$ is constant as a function of $z$ ("simplifying assumption"), then
$$C_{X,Y;Z}(x, y) = C_{X,Y|Z}(x, y).$$
\end{corollary}

\begin{proof}
This result is a direct consequence of integrating $C_{X,Y|Z=z}(x, y) f_Z(z)$ with respect to $z$ and taking into account that, from the theorem assumption, $C_{X,Y|Z=z}(x, y)$ is constant as a function of $z$.
\begin{align*}
C_{X,Y;Z}(x, y) &= \int_{-\infty}^{\infty} C_{X,Y|Z=z}(x, y) f_Z(z) dz \\
&= \int_{-\infty}^{\infty} C_{X,Y|Z}(x, y) f_Z(z) dz \\
&= C_{X,Y|Z}(x, y) \int_{-\infty}^{\infty} f_Z(z) dz \\
&= C_{X,Y|Z}(x, y).
\end{align*}
\end{proof}

From this result, it also becomes clear that when $(X, Y)$ are conditionally independent given $Z$ (which produces the conditional copula $C_{X,Y|Z=z}(x, y) = xy$), then the partial copula is the independence copula, which is coherent with lemma 3.1 of @patra2016. However, the reciprocal statement is not true - see next remark.

\begin{remark}\label{rem:partial-copula-necessary-not-sufficient}
Corollary \ref{cor:partial-conditional-copula} entails that, if $X \independent Y | Z$, then $C_{X,Y;Z}(x, y) = xy$ (or, equivalently, $\UX$ and $\UY$ are independent).

On the other hand, if $\UX$ and $\UY$ are independent, it is not necessarily true that $X \independent Y | Z$. Consider the following example\footnote{It is interesting noticing that, if $X \independent Z$, $Y \independent Z$ and $X,Y \sim U(0,1)$, then this example produces a trivariate Farlie-Gumbel-Morgenstern copula that commonly appears in the literature as a counterexample for certain statistical dependence facts \citep{nelsen2012, gonzalezlopez2024}}: if $Z \sim U(0,1)$ and $C_{X,Y|Z=z}(x, y) = xy + (2z - 1)xy(1-x)(1-y)$ (that is, a Farlie-Gumbel-Morgenstern copula with parameter $2z - 1$), then from Proposition \ref{prop:partial-copula},
$$C_{X,Y;Z}(x, y) = \int_0^1 [xy + (1-2z)xy(1-x)(1-y)] dz = xy,$$
which is the independence copula, which entails, from the definition of the partial copula, that $\UX$ and $\UY$ are independent. However, except for $Z = 0.5$, $X$ and $Y$ are not conditionally independent given $Z = z$.

Therefore, $\UX \independent \UY$ is a necessary but not sufficient condition for $X \independent Y | Z$. A possible explanation for this fact is that, in the example presented in the remark, the conditional association of $(X, Y)$ given $Z = z$ is positive for some values of $z$ and negative for other values of $z$. In this sense, the partial copula represents a kind of "average conditional association"; this fact is shown in next remark.
\end{remark}

\begin{remark}\label{rem:partial-copula-expected-conditional-copula}
Let $g(x, y, z) = C_{X,Y|Z=z}(x, y)$. $g(x, y, Z)$ is a random variable, and Proposition \ref{prop:partial-copula} entails that
$$C_{X,Y;Z}(x, y) = E[g(x, y, Z)],$$
which means that the partial copula is the expected conditional copula.
\end{remark}

## Partial copula as an extension of partial correlation

Next remark underscores the connection between the partial copula and the notion of partial correlation.

\begin{remark}\label{rem:partial-correlation}
A possible way to define the partial correlation between two random variables $X$ and $Y$, given $Z$, is to construct two regression models $X = \alpha + \beta Z + \epsilon_{X;Z}$ and $Y = \gamma + \theta Z + \epsilon_{Y;Z}$ and them compute the Pearson's correlation between the error terms $\epsilon_{X;Z}$ and $\epsilon_{Y;Z}$ \citep{kim2015}.

More recently, \cite{patra2016} studied the problem of how to construct a nonparametric notion of error term, that is, a notion of error that does not assume a specific functional form for the relationship between the regressor and the response variable. Imposing some conditions that are reasonable to expect from a notion of error term (such as being independent fro the regressor), \cite{patra2016} showed that the Rosenblatt transforms $F_{X|Z}(X|Z)$ and $F_{Y|Z}(Y|Z)$ satisfy those conditions and, therefore, may be seen as a general notion of error term. \citep[for further considerations, see also Section 4 of][]{justus2024}

Partial copulas are the joint CDF of the nonparametric error terms $F_{X|Z}(X|Z)$ and $F_{Y|Z}(Y|Z)$, and, therefore, can be seen as a nonparametric extension of the notion of partial correlation.
\end{remark}

Previous results in the literate have established that zero partial correlation is not equivalent to conditional independence, except in some special cases such as the multivariate normal distribution [@baba2004]. Next example shows that, even under conditional independence, the partial correlation may be arbitrarily close to 1, which suggests that the partial correlation may be a very poor measure of covariate-adjusted association.

\begin{example}\label{ex:partial-correlation}
Let be $Z \sim N(0,1)$. Let $X|Z=z \sim N(z^2, \sigma^2)$ and $Y|Z=z \sim N(z^2, \sigma^2)$ be conditionally independent given $Z=z$. It is straightforward to see that, in this case,
$$Corr(X,Z) = Corr(Y,Z) = 0,$$
which entails, from the definition of partial correlation, that $Corr(X,Y;Z) = Corr(X,Y)$. On the other hand, $X$ and $Y$ are exchangeable and, from Theorem 2 in \cite{oneill2009}, it holds that
$$Corr(X,Y) =  \frac{Var(E(X|Z))}{Var(E(X|Z)) + E(Var(X|Z))} = \frac{Var(Z^2)}{Var(Z^2) + \sigma^2}.$$
Therefore,
$$Corr(X,Y;Z) = Corr(X,Y) = \frac{2}{2 + \sigma^2} > 0,$$
which becomes arbitrarily close to 1 as $\sigma^2$ approaches to 0, even though $X$ and $Y$ are conditionally independent given $Z$.
\end{example}

The result seen in Example \ref{ex:partial-correlation} may look worrying: even when two variables are conditionally independent given a covariate, their partial correlation may be close to 1. An intuitive explanation for this fact may comes from Remark \ref{rem:partial-correlation}: given that the partial correlation is the Pearson's correlation between the error terms of two linear regression models, only the "linear effect" of $Z$ on the variables is being "removed". In Example \ref{ex:partial-correlation}, the expectation of each variable given $Z$ is not linear (not even monotonic), and, therefore, the partial correlation fails in removing the effect of $Z$ on the variables. On the other hand, if  $X$ and $Y$ are conditionally independent given $Z$, then the partial copula is always the independence copula. In this sense, the partial copula fully remove the "spurious association" induced by $Z$. Therefore, except when a researcher has strong reasons to make some parametric assumptions (such as linearity), the partial copula looks to be a much more appropriate notion for representing covariate-adjusted statistical association.

## Effect of properties of the conditional copulas on the partial copula

Given the functional relationship between the partial and the conditional copulas, an interesting research line that emerges is studying in what extent properties of the conditional copulas constrain the properties of the partial copula. A first result on this matter is given by next proposition. The result is based on the notions of quadrant positive and quadrant negative dependence: a copula $C$ is said to have quadrant positive dependence (QPD) if $C(u, v) \geq uv$ for all $(u, v) \in [0,1]^2$. In an analogous way, the copula has quadrant negative dependence (QND) if $C(u, v) \leq uv$ for all $(u, v) \in [0,1]^2$ [@joe2014]. A direct consequence of the fact that the partial copula is an expected conditional copula is given by the following proposition.

\begin{proposition}\label{prop:quadrant-dependence}
Let $(X, Y, Z)$ be a continuous random vector. If the conditional copula of $(X, Y)$ given $Z = z$ has positive quadrant dependence for all $z \in \mathbb{R}$, then the partial copula $C_{X,Y;Z}(x, y)$ has positive quadrant dependence.

In the same way, if the conditional copula of $(X, Y)$ given $Z = z$ has negative quadrant dependence for all $z \in \mathbb{R}$, then the partial copula $C_{X,Y;Z}(x, y)$ has negative quadrant dependence.
\end{proposition}

\begin{proof}
If $g(x, y, z) = C_{X,Y|Z=z}(x, y) \geq xy$ for all $z \in \mathbb{R}$, then, by interpreting the constant $xy$ as a degenerate random variable and applying the monotonicity of expectation \citep[property E2 of section 3.3 of][]{james1981}, we have the inequality
$$xy \leq E[g(x, y, Z)] = C_{X,Y;Z}(x, y).$$
Therefore, $C_{X,Y;Z}(x, y) \geq xy$, as we wished to prove. The same argument holds for negative quadrant dependence.
\end{proof}

An intuitive interpretation for Proposition \ref{prop:quadrant-dependence} is that, whenever the conditional association between $X$ and $Y$ is positive [negative] for all values of $Z$, the partial copula also exhibits positive [negative] dependence.

Next proposition uses a dependence measure based on Kolmogorov distance, which, for simplicity, we will call it as Kolmogorov distance dependence (KDD). This coefficient was introduced in @wolff1977 and @schweizer1981, and is defined as
$$D(X, Y) = 4 \sup_{x,y \in [0,1]} (|C_{X,Y}(x, y) - xy|).$$
It is immediate to see that KDD is proportional to the Kolmogorov distance between a given copula and the independence copula; the constant 4 ensures that $D(X, Y) \in [0, 1]$. This coefficient is constructed intending to measure in what extent a copula deviates from independence, without distinguishing between "positive" and "negative" dependence; in this sense, values closer to 0 represent independence and cases closer to 1 represent strong dependencee - for further details, see @wolff1977 and @schweizer1981^[Note that the behavior of KDD differs from the behavior of the so-called "concordance measures", such as Separman's
correlation coefficient, which may be equal to zero even if the two variables are not independent [@garciaportugues2025].
This happens intuitively because dependence and concordance measures have two different goals: the former ones measure
deviation from indepenendence, and the last ones measures the degree os monothonic association [see @nelsen2006; @joe2014].]. The following proposition shows how the value of KDD in the conditional copulas
constrains the value of KDD in the partial copula.

\begin{proposition}\label{prop:partial-copula-independence}
Let $X, Y, Z$ be jointly continuous random variables. Then,
$$D(\UX, \UY) \leq \sup_{z \in \mathbb{R}} (D(X, Y | Z = z)).$$
\end{proposition}

\begin{proof}
Suppose $\sup_{z \in \mathbb{R}}(D(X, Y | Z = z)) = k$, for some $k \in [0,1]$. This means that, for every $z \in \mathbb{R}$, it holds that
$$k \geq D(X, Y | Z = z) = 4 \sup_{x,y \in [0,1]} (|C_{X,Y|Z=z}(x, y) - xy|),$$
which entails that $|C_{X,Y|Z=z}(x, y) - xy| \leq k/4$ for every $x, y \in [0, 1]$, $z \in \mathbb{R}$. Therefore,
$$C_{X,Y|Z=z}(x, y) \in \left[xy - \frac{k}{4}, xy + \frac{k}{4}\right].$$

The remainder of the proof comes from remembering that the partial copula evaluated in a point $(x, y) \in [0, 1]^2$ is the expectation of a random variable (see Remark \ref{rem:partial-copula-expected-conditional-copula}), which enables using the monotonicity of expectation: property E2 of section 3.3 of \citet{james1981} readily entails that, if a random variable belongs almost surely to an interval, then the same happens to its expectation. Therefore,
$$xy - \frac{k}{4} \leq C_{X,Y;Z}(x, y) \leq xy + \frac{k}{4},$$
for every $x, y \in [0, 1]$. Therefore, $|C_{X,Y;Z}(x, y) - xy| \leq k/4$, which entails that
$$D(\UX, \UY) = 4 \sup_{x,y \in [0,1]} (|C_{X,Y}(x, y) - xy|) \leq k.$$
\end{proof}

The last result can be intuitively interpreted in the following way: regardless of simplifying assumption holding or not, if all the conditional copulas are distant from independence at most $k$, then the partial copula is distant from independence at most $k$ too. This entails that, if all conditional copulas are close to the independence case, then the partial copula is necessarily close to independence.

Furthermore, it is possible to also bound Spearman's correlation coefficient (here denoted by letter $\rho$) in the partial copula, provided an upper bound for the KDD of the conditional copulas is also known (it is worth remembering that Spearman's correlation is also a functional of the copula, see @fredricks2007).

\begin{proposition}\label{def:spearman-bounds}
Let $X, Y, Z$ be jointly continuous random variables. If $(D(X, Y | Z = z)) \leq k$ for all $z \in \mathbb{R}$, then
$$\max(-1, -3k) \leq \rho(X, Y; Z) \leq \min(1, 3k).$$
\end{proposition}

\begin{proof}
The proof of Proposition \ref{prop:partial-copula-independence} entails that, if $(D(X, Y | Z = z)) \leq k$ for all $z \in \mathbb{R}$, then
$$xy - \frac{k}{4} \leq C_{X,Y;Z}(x, y) \leq xy + \frac{k}{4},$$
which entails that
$$\int_0^1 \int_0^1 C_{X,Y;Z}(x, y) dx dy \in \left[\frac{1-k}{4}, \frac{1+k}{4}\right].$$

Theorem 5.1.6 in \citet{nelsen2006} states that, given a copula $C$, its Spearman's correlation can be represented as $12 \int_0^1 \int_0^1 C_{X,Y;Z}(x, y) dx dy - 3$. Applying this representation in the partial copula, it holds that
$$-3k \leq \rho(X, Y; Z) \leq 3k,$$
as we wished to prove. The final bound in the proposition is obtained taking into account that Spearman's coefficient belongs to $[-1, 1]$ range \citep[see Theorem 5.1.9 in][]{nelsen2006}.
\end{proof}

An even better bound can be obtained for the Kendall's tau correlation coefficient (denoted by $\tau$), which is also a functional of the copula [@nelsen2006] - see next Proposition.

\begin{proposition}
Let $X, Y, Z$ be jointly continuous random variables. If $(D(X, Y | Z = z)) \leq k$ for all $z \in \mathbb{R}$, then
$$\max(-1, -2k) \leq \tau(X, Y; Z) \leq \min(1, 2k).$$
\end{proposition}

\begin{proof}
We have seen in the proof of Proposition \ref{prop:partial-copula-independence} that, if $(D(X, Y | Z = z)) \leq k$ for all $z \in \mathbb{R}$, then
$$xy - \frac{k}{4} \leq C_{X,Y;Z}(x, y) \leq xy + \frac{k}{4}.$$
Therefore, inserting $U_X$ and $U_Y$ as arguments, we have
$$U_X U_Y - \frac{k}{4} \leq C_{X,Y;Z}(U_X, U_Y) \leq U_X U_Y + \frac{k}{4}.$$
The monotonicity of expectation \citep[property E2 of section 3.3 of][]{james1981} entails that
$$E[U_X U_Y] - \frac{k}{4} \leq E[C_{X,Y;Z}(U_X, U_Y)] \leq E[U_X U_Y] + \frac{k}{4}.$$
The Equation 5.1.15b from Theorem 5.1.6 in \citet{nelsen2006} entails that $\rho(X,Y;Z) = 12 E[U_X U_Y] -3$, which means that
$$\frac{\rho(X,Y;Z) + 3}{12} - \frac{k}{4} \leq E[C_{X,Y;Z}(U_X, U_Y)] \leq \frac{\rho(X,Y;Z) + 3}{12} + \frac{k}{4}.$$
The final argument comes from remembering that, according to Theorem 5.1.3 in \citet{nelsen2006}, Kendall's tau correlation coefficient can be represented as $4E[C_{X,Y;Z}(U_X, U_Y)] - 1$, which leads us to the conclusion that
$$\frac{\rho(X,Y;Z) + 3}{3} - k -1 \leq \tau(X, Y; Z) \leq \frac{\rho(X,Y;Z) + 3}{3} + k -1,$$
and, from the inequality $-3k \leq \rho(X, Y; Z) \leq 3k,$ it follows that
$$-2k \leq \tau(X, Y; Z) \leq 2k.$$
As in the previous proof, the final bounds are obtained using the fact that Kendall's coefficient belongs to the $[-1, 1]$ range \citep[see Theorem 5.1.9 in][]{nelsen2006}.
\end{proof}

Additionally, the connection between the correlation in the conditional copulas and in the partial copula can be summarized in the following theorem.

\begin{proposition}\label{prop:spearman-partial-conditional}
Let $X, Y, Z$ be jointly continuous random variables. Then,
$$\rho(X, Y; Z) = \int_{-\infty}^{\infty} \rho(X, Y | Z = z) f_Z(z) dz.$$
\end{proposition}

\begin{proof}
Applying Equation 5.1.15c of \citet{nelsen2006} in the partial copula, we obtain
$$\rho(X, Y; Z) = 12 \int_0^1 \int_0^1 C_{X,Y;Z}(x, y) dx dy - 3.$$
Now using representation from Proposition \ref{prop:partial-copula}, we obtain
\begin{align*}
\rho(X, Y; Z) &= 12 \int_0^1 \int_0^1 \int_{-\infty}^{\infty} C_{X,Y|Z=z}(x, y) f_Z(z) dz dx dy - 3 \\
&= 12 \int_{-\infty}^{\infty} \left(\int_0^1 \int_0^1 C_{X,Y|Z=z}(x, y) dx dy \right) f_Z(z) dz - 3 \\
&= 12 \int_{-\infty}^{\infty} \left(\frac{\rho(X, Y | Z = z) + 3}{12}\right) f_Z(z) dz - 3 \\
&= \int_{-\infty}^{\infty} \rho(X, Y | Z = z) f_Z(z) dz.
\end{align*}
\end{proof}

That is, the Spearman's correlation induced by the partial copula is the expected conditional Spearman's correlation - this greatly clarifies the role of conditional correlations in determining partial correlation. Finally, a consequence of this result in a family of copulas studied in @rodriguezlallena2004 is illustrated in the following example.

\begin{example}
\cite{rodriguezlallena2004} studied a nonparametric of copulas where the copula of two random variables $X',Y'$ can be written as $C_{X',Y'}(x, y) = xy + f(x)g(y)$. Now, consider a case where all the conditional copulas of $X,Y$ given $Z$ can be written as
$$C_{X,Y|Z=z}(x, y) = xy + f(x,z)g(y,z),$$
which means that all the conditional copulas belong to the family of copulas studied in \cite{rodriguezlallena2004}. \cite{komelj2010} showed that all copulas on this family have Spearman's correlation coefficient belonging to $[-3/4, 3/4]$ Therefore, applying Proposition \ref{prop:spearman-partial-conditional} and monotonicity of expectation \citep[property E2 of section 3.3 of][]{james1981}, is straightforward to see that
$$-\frac{3}{4} \leq \rho(X, Y; Z) \leq \frac{3}{4}.$$
Therefore, all the conditional copulas belonging to the family of copulas studied in \cite{rodriguezlallena2004} entails that the Spearman's correlation coefficient in the partial copula is also bounded by $[-3/4, 3/4]$.
\end{example}

# Simulation study

```{r simulation, echo=FALSE}
library(rvinecopulib); library(ggplot2)
library(dplyr); library(tidyr)

set.seed(123); n <- 5000

structure <- cvine_structure(c(1, 2, 3))
indep_par <- numeric(0)

# core sim function
simulate_vine <- function(family_xz, rot_xz, par_xz, 
                          family_yz, rot_yz, par_yz,
                          family_xy_z, rot_xy_z, par_xy_z) {
  
  bicop_xz <- bicop_dist(family_xz, rot_xz, par_xz)
  bicop_yz <- bicop_dist(family_yz, rot_yz, par_yz)
  bicop_xy_z <- bicop_dist(family_xy_z, rot_xy_z, par_xy_z)
  
  vine <- vinecop_dist(list(list(bicop_xz, bicop_yz), list(bicop_xy_z)), structure)
  data <- rvinecop(n, vine)
  
  X <- data[, 1]; Y <- data[, 2]; Z <- data[, 3]
  U_X <- hbicop(cbind(X, Z), cond_var = 2, bicop_xz)
  U_Y <- hbicop(cbind(Y, Z), cond_var = 2, bicop_yz)
  
  data.frame(X = X, Y = Y, Z = Z, U_X = U_X, U_Y = U_Y)
}

# scenario configurations (1-10)
configs <- list(
  list(scen = 1, fam = "frank",   pars = list(c(1,1,1), c(3,3,1), c(5,5,1))),
  list(scen = 1, fam = "gumbel",  pars = list(c(1.5,1.5,1.5), c(2,2,1.5), c(3,3,1.5))),
  list(scen = 1, fam = "clayton", pars = list(c(1,1,1), c(3,3,1), c(5,5,1))),
  
  list(scen = 2, fam = "frank",   pars = list(c(1,1,NA), c(3,3,NA), c(5,5,NA)), cond_indep = TRUE),
  list(scen = 2, fam = "gumbel",  pars = list(c(1.5,1.5,NA), c(2,2,NA), c(3,3,NA)), cond_indep = TRUE),
  list(scen = 2, fam = "clayton", pars = list(c(1,1,NA), c(3,3,NA), c(5,5,NA)), cond_indep = TRUE),
  
  list(scen = 3, fam = "frank",      pars = list(c(-1,-1,NA), c(-3,-3,NA), c(-5,-5,NA)), cond_indep = TRUE),
  list(scen = 3, fam = "gumbel_90",  pars = list(c(1.5,1.5,NA), c(2,2,NA), c(3,3,NA)), rot = 90, cond_indep = TRUE),
  list(scen = 3, fam = "clayton_90", pars = list(c(1,1,NA), c(3,3,NA), c(5,5,NA)), rot = 90, cond_indep = TRUE),
  
  list(scen = 4, fam = "frank",    pars = list(c(1,-1,NA), c(3,-3,NA), c(5,-5,NA)), cond_indep = TRUE),
  list(scen = 4, fam = "gaussian", pars = list(c(0.3,-0.3,NA), c(0.5,-0.5,NA), c(0.7,-0.7,NA)), cond_indep = TRUE),
  
  list(scen = 5, fam = "frank",    pars = list(c(1,1,-1), c(1,1,-3), c(1,1,-5))),
  list(scen = 5, fam = "gaussian", pars = list(c(0.2,0.2,-0.2), c(0.2,0.2,-0.5), c(0.2,0.2,-0.7))),
  list(scen = 5, fam = "frank",    pars = list(c(5,5,-1), c(5,5,-0.5))),
  list(scen = 5, fam = "gaussian", pars = list(c(0.7,0.7,-0.2), c(0.7,0.7,-0.1))),
  
  list(scen = 6, fam = "frank",      pars = list(c(-1,-1,1), c(-3,-3,1), c(-5,-5,1))),
  list(scen = 6, fam = "gumbel_90",  pars = list(c(1.5,1.5,1.5), c(2,2,1.5), c(3,3,1.5)), rot_xz = 90, rot_yz = 90),
  list(scen = 6, fam = "clayton_90", pars = list(c(1,1,1), c(3,3,1), c(5,5,1)), rot_xz = 90, rot_yz = 90),
  
  list(scen = 7, fam = "frank",      pars = list(c(-1,-1,-1), c(-1,-1,-3), c(-1,-1,-5))),
  list(scen = 7, fam = "gumbel_90",  pars = list(c(1.5,1.5,1.5), c(1.5,1.5,2), c(1.5,1.5,3)), rot = 90),
  list(scen = 7, fam = "clayton_90", pars = list(c(1,1,1), c(1,1,3), c(1,1,5)), rot = 90),
  
  list(scen = 8, fam = "frank",    pars = list(c(1,-1,1), c(3,-3,1), c(5,-5,1))),
  list(scen = 8, fam = "gaussian", pars = list(c(0.2,-0.2,0.2), c(0.5,-0.5,0.2), c(0.7,-0.7,0.2))),
  
  list(scen = 9, fam = "frank",    pars = list(c(1,-1,-1), c(3,-3,-1), c(5,-5,-1))),
  list(scen = 9, fam = "gaussian", pars = list(c(0.2,-0.2,-0.2), c(0.5,-0.5,-0.2), c(0.7,-0.7,-0.2))),
  
  list(scen = 10, fam = "frank", pars = list(c(NA,NA,-1), c(NA,NA,1)), marginal_indep = TRUE),
  list(scen = 10, fam = "indep", pars = list(c(NA,NA,NA)), marginal_indep = TRUE, cond_indep = TRUE)
)

# run all simulations (scenarios 1-10)
all_data <- data.frame()
results <- data.frame()
case_counter <- list()

for (cfg in configs) {
  base_fam <- gsub("_90", "", cfg$fam)
  rot <- if (!is.null(cfg$rot)) cfg$rot else 0
  rot_xz <- if (!is.null(cfg$rot_xz)) cfg$rot_xz else rot
  rot_yz <- if (!is.null(cfg$rot_yz)) cfg$rot_yz else rot
  
  for (i in seq_along(cfg$pars)) {
    p <- cfg$pars[[i]]
    
    if (isTRUE(cfg$marginal_indep)) {
      fam_xz <- "indep"; par_xz <- indep_par; r_xz <- 0
      fam_yz <- "indep"; par_yz <- indep_par; r_yz <- 0
    } else {
      fam_xz <- base_fam; par_xz <- p[1]; r_xz <- rot_xz
      fam_yz <- base_fam; par_yz <- p[2]; r_yz <- rot_yz
    }
    
    if (isTRUE(cfg$cond_indep)) {
      fam_xy_z <- "indep"; par_xy_z <- indep_par; r_xy_z <- 0
    } else {
      fam_xy_z <- base_fam; par_xy_z <- p[3]; r_xy_z <- rot
    }
    
    df <- simulate_vine(fam_xz, r_xz, par_xz, fam_yz, r_yz, par_yz, fam_xy_z, r_xy_z, par_xy_z)
    
    scen_key <- as.character(cfg$scen)
    if (is.null(case_counter[[scen_key]])) case_counter[[scen_key]] <- 0
    case_counter[[scen_key]] <- case_counter[[scen_key]] + 1
    case_letter <- letters[case_counter[[scen_key]]]
    case_suffix <- ""
    
    clean_fam <- gsub("_90", "", cfg$fam)
    display_fam <- paste0(toupper(substr(clean_fam, 1, 1)), substr(clean_fam, 2, nchar(clean_fam)))
    rot_label <- if (grepl("_90", cfg$fam)) " (R)" else ""
    par_xz_str <- ifelse(is.na(p[1]), "ind", p[1])
    par_yz_str <- ifelse(is.na(p[2]), "ind", p[2])
    par_xy_z_str <- ifelse(isTRUE(cfg$cond_indep), "ind", p[3])
    label <- sprintf("%s%s, θ = (%s, %s, %s)", display_fam, rot_label, par_xz_str, par_yz_str, par_xy_z_str)
    
    rho_xy <- cor(df$X, df$Y, method = "spearman")
    rho_uv <- cor(df$U_X, df$U_Y, method = "spearman")
    tau_xy <- cor(df$X, df$Y, method = "kendall")
    tau_uv <- cor(df$U_X, df$U_Y, method = "kendall")
    
    case_id <- paste0(scen_key, case_letter, case_suffix)
    
    df$Scenario <- cfg$scen
    df$Case <- case_id
    df$Label <- label
    df$Case_Label <- paste0(case_letter, case_suffix, ": ", label)
    df$Family <- display_fam
    df$rho_XY <- rho_xy
    df$rho_UV <- rho_uv
    df$tau_XY <- tau_xy
    df$tau_UV <- tau_uv
    
    all_data <- rbind(all_data, df)
    
    results <- rbind(results, data.frame(
      Scenario = cfg$scen,
      Case = case_id,
      Family = display_fam,
      Parameters = label,
      rho_XY = round(rho_xy, 3),
      rho_UV = round(rho_uv, 3),
      tau_XY = round(tau_xy, 3),
      tau_UV = round(tau_uv, 3)
    ))
  }
}

# scenario 11: violation of the simplifying assumption
simulate_nonsimplified <- function(par_func, family = "frank") {
  bicop_xz <- bicop_dist("frank", 0, 1)
  bicop_yz <- bicop_dist("frank", 0, 1)
  
  Z <- runif(n)
  X <- Y <- U_X <- U_Y <- numeric(n)
  
  for (i in 1:n) {
    param_z <- par_func(Z[i])
    cond_cop <- bicop_dist(family, 0, param_z)
    ST <- rbicop(1, cond_cop)
    X[i] <- hbicop(c(ST[1,1], Z[i]), cond_var = 2, bicop_xz, inverse = TRUE)
    Y[i] <- hbicop(c(ST[1,2], Z[i]), cond_var = 2, bicop_yz, inverse = TRUE)
    U_X[i] <- hbicop(c(X[i], Z[i]), cond_var = 2, bicop_xz)
    U_Y[i] <- hbicop(c(Y[i], Z[i]), cond_var = 2, bicop_yz)
  }
  
  data.frame(X = X, Y = Y, Z = Z, U_X = U_X, U_Y = U_Y)
}

nonsimplified_configs <- list(
  list(case = "11a", family = "frank",    par_func = function(z) exp(z),  label = "Frank, θ(z) = exp(z)"),
  list(case = "11b", family = "frank",    par_func = function(z) -exp(z), label = "Frank, θ(z) = -exp(z)"),
  list(case = "11c", family = "gaussian", par_func = function(z) 1 - 2*z, label = "Gaussian, θ(z) = 1-2z")
)

data_11 <- data.frame()
results_11 <- data.frame()

for (cfg in nonsimplified_configs) {
  df <- simulate_nonsimplified(cfg$par_func, cfg$family)
  
  rho_xy <- cor(df$X, df$Y, method = "spearman")
  rho_uv <- cor(df$U_X, df$U_Y, method = "spearman")
  tau_xy <- cor(df$X, df$Y, method = "kendall")
  tau_uv <- cor(df$U_X, df$U_Y, method = "kendall")
  
  df$Scenario <- 11
  df$Case <- cfg$case
  df$Label <- cfg$label
  case_letter <- substr(cfg$case, nchar(cfg$case), nchar(cfg$case))
  df$Case_Label <- paste0(case_letter, ": ", cfg$label)
  df$Family <- cfg$family
  df$rho_XY <- rho_xy
  df$rho_UV <- rho_uv
  df$tau_XY <- tau_xy
  df$tau_UV <- tau_uv
  
  data_11 <- rbind(data_11, df)
  
  results_11 <- rbind(results_11, data.frame(
    Scenario = 11,
    Case = cfg$case,
    Family = cfg$family,
    Parameters = cfg$label,
    rho_XY = round(rho_xy, 3),
    rho_UV = round(rho_uv, 3),
    tau_XY = round(tau_xy, 3),
    tau_UV = round(tau_uv, 3)
  ))
}

# combine all data
all_data <- rbind(all_data, data_11)
results <- rbind(results, results_11)

scenario_info <- data.frame(
  Scenario = 1:11,
  Title = c(
    "Scenario 1: Positive Confounding, Positive Conditional (+,+,+)",
    "Scenario 2: Positive Confounding, Conditional Independence (+,+,ind)",
    "Scenario 3: Negative Confounding, Conditional Independence (-,-,ind)",
    "Scenario 4: Opposite Confounding, Conditional Independence (+,-,ind)",
    "Scenario 5: Positive Confounding, Negative Conditional (+,+,-)",
    "Scenario 6: Negative Confounding, Positive Conditional (-,-,+)",
    "Scenario 7: Negative Confounding, Negative Conditional (-,-,-)",
    "Scenario 8: Opposite Confounding, Positive Conditional (+,-,+)",
    "Scenario 9: Opposite Confounding, Negative Conditional (+,-,-)",
    "Scenario 10: No Confounding (ind,ind,*)",
    "Scenario 11: Simplifying Assumption Violated"
  ),
  Short = c("1: (+,+,+)", "2: (+,+,ind)", "3: (-,-,ind)", "4: (+,-,ind)", "5: (+,+,-)",
            "6: (-,-,+)", "7: (-,-,-)", "8: (+,-,+)", "9: (+,-,-)", "10: (ind,ind,*)",
            "11: Non-simplified")
)

```

The theoretical results in Section 3 establish that the partial copula $C_{X,Y;Z}$ captures the dependence between $X$ and $Y$ after removing the effect of $Z$. In particular, Proposition \ref{prop:spearman-partial-conditional} shows that the Spearman correlation induced by the partial copula equals the expected conditional correlation, while Proposition \ref{prop:quadrant-dependence} establishes that the sign of quadrant dependence is preserved when it is constant across all conditional copulas. In this section, we conduct a simulation study to illustrate these results and to demonstrate how the partial copula can reveal the true conditional association structure when marginal correlations are confounded. All simulations were performed in R using the \texttt{rvinecopulib} package [@rvinecopulib]. The manuscript and all analyses are fully reproducible through the \texttt{rix} package [@rix], and the source code and environment files are available in the accompanying repository.

## Design

We employ a C-vine copula with $Z$ as the root node, which decomposes the joint distribution of $(X, Y, Z)$ through the pair copulas $C_{X,Z}$, $C_{Y,Z}$, and $C_{X,Y|Z}$. For each configuration, we generate a single sample of $n = 5000$ observations $(X_i, Y_i, Z_i)_{i=1}^{n}$ and compute the partial copula pseudo-observations via h-functions,
$$U_{X,i} = h_2(X_i, Z_i; C_{X,Z}) \quad \text{and} \quad U_{Y,i} = h_2(Y_i, Z_i; C_{Y,Z}),$$
where $h_2(u, v; C) = \partial C(u,v)/\partial v$ is the conditional distribution function of the first argument given the second.

To assess how confounding and conditional dependence interact, we compare the marginal and partial rank correlations. Specifically, we compute Spearman's $\rho$ and Kendall's $\tau$ for both the original pair $(X, Y)$ and the partial copula pair $(U_X, U_Y)$. A discrepancy between these quantities indicates confounding: dependence between $X$ and $Y$ that is induced by $Z$ rather than by direct conditional association.

We consider eleven scenarios, summarized in Table 1. The triplet $(s_{XZ}, s_{YZ}, s_{XY|Z})$ indicates the sign of dependence in each pair copula: $+$ for positive, $-$ for negative, and $\text{ind}$ for independence. Scenarios 1--10 satisfy the simplifying assumption, so that $C_{X,Y|Z=z}$ does not vary with $z$ and Corollary \ref{cor:partial-conditional-copula} applies. Scenario 11 violates the simplifying assumption by letting the conditional copula parameter vary as a function of $Z$ through specifications $\theta(z) = \exp(z)$, $\theta(z) = -\exp(z)$, and $\theta(z) = 1 - 2z$; the last of these illustrates Remark \ref{rem:partial-copula-necessary-not-sufficient}, where positive and negative conditional associations cancel out in expectation.

Across scenarios, we consider multiple copula families (Frank, Gumbel, Clayton, and Gaussian) over a range of parameter values to ensure robustness of the findings. Negative dependence for Gumbel and Clayton copulas is achieved via 90-degree rotation.

\small

| Scenario | Structure | Description |
|:--------:|:---------:|:------------|
| 1 | $(+,+,+)$ | Positive confounding, positive conditional |
| 2 | $(+,+,\text{ind})$ | Positive confounding, conditional indep. |
| 3 | $(-,-,\text{ind})$ | Negative confounding, conditional indep. |
| 4 | $(+,-,\text{ind})$ | Opposite confounding, conditional indep. |
| 5 | $(+,+,-)$ | Positive confounding, negative conditional |
| 6 | $(-,-,+)$ | Negative confounding, positive conditional |
| 7 | $(-,-,-)$ | Negative confounding, negative conditional |
| 8 | $(+,-,+)$ | Opposite confounding, positive conditional |
| 9 | $(+,-,-)$ | Opposite confounding, negative conditional |
| 10 | $(\text{ind},\text{ind},\cdot)$ | No confounding |
| 11 | --- | Simplifying assumption violated |

Table: Summary of simulation scenarios. The triplet $(s_{XZ}, s_{YZ}, s_{XY|Z})$ denotes the sign of dependence in each pair copula: $+$ positive, $-$ negative, $\text{ind}$ independence, and $\cdot$ varies across configurations.

\normalsize

## Results

The scatter plots in the Appendix display Spearman's $\rho$ and Kendall's $\tau$ for both the marginal pair $(X,Y)$ and the partial copula pair $(U_X, U_Y)$ across all scenarios. The results confirm the theoretical predictions. In conditional independence scenarios (2, 3, and 4), the partial correlations are negligible while marginal correlations can be substantial, demonstrating that the partial copula successfully removes confounding. In scenarios where confounding reinforces conditional dependence (1, 6, and 9), the partial correlations correctly recover the conditional association, which is attenuated relative to the marginal correlations. Notably, scenarios 5, 7, and 8 exhibit Simpson's paradox: marginal and partial correlations have opposite signs. In these scenarios, the confounding effect opposes the conditional dependence --- for instance, in scenario 7 $(-,-,-)$, the product of two negative marginal dependencies induces a positive spurious correlation that counteracts the negative conditional dependence. This illustrates how the partial copula can reveal conditional associations that are masked or reversed by confounding. Finally, Scenario 11 examines violations of the simplifying assumption. In cases 11a and 11b, where $\theta(z) = \exp(z)$ and $\theta(z) = -\exp(z)$ respectively, the sign of conditional dependence remains constant across all values of $Z$, and the partial correlations closely match the marginal correlations, consistent with Proposition \ref{prop:quadrant-dependence}. In contrast, case 11c deserves particular attention. Here the conditional copula parameter $\theta(z) = 1 - 2z$ is a Gaussian copula whose correlation switches sign at $z = 0.5$: for $z < 0.5$ the conditional dependence is positive, while for $z > 0.5$ it becomes negative. As a result, the partial copula --- which averages over all conditional copulas --- yields near-zero partial correlation despite strong conditional dependence for almost all values of $Z$, confirming Remark \ref{rem:remark-2}. Figure \ref{fig:scenario-11c-detail} illustrates this by displaying the conditional scatter plots for $Z < 0.5$ and $Z > 0.5$ alongside the partial copula: the opposing dependence structures are clearly visible in the conditional panels but cancel out in the partial copula. This highlights a fundamental limitation: the partial copula captures average conditional dependence and may not be the best tool when one is interested in local, $z$-specific dependence.


```{r scenario-11c-detail, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.height=4, fig.width=10, fig.cap="Scenario 11c: Gaussian copula with $\\theta(z) = 1 - 2z$. Left: conditional scatter for $Z < 0.5$ (positive dependence). Center: conditional scatter for $Z > 0.5$ (negative dependence). Right: partial copula averaging over all $Z$. Color gradient represents $Z$; $\\rho$ denotes Spearman's correlation."}
df_11c <- all_data %>% filter(Case == "11c")

df_low  <- df_11c %>% filter(Z < 0.5)
df_high <- df_11c %>% filter(Z > 0.5)

rho_low  <- round(cor(df_low$U_X, df_low$U_Y, method = "spearman"), 2)
rho_high <- round(cor(df_high$U_X, df_high$U_Y, method = "spearman"), 2)
rho_all  <- round(cor(df_11c$U_X, df_11c$U_Y, method = "spearman"), 2)

plot_data_11c <- rbind(
  df_low  %>% mutate(Panel = sprintf("Conditional: Z < 0.5 (ρ = %s)", rho_low)),
  df_high %>% mutate(Panel = sprintf("Conditional: Z > 0.5 (ρ = %s)", rho_high)),
  df_11c  %>% mutate(Panel = sprintf("Partial copula (ρ = %s)", rho_all))
) %>%
  mutate(Panel = factor(Panel, levels = c(
    sprintf("Conditional: Z < 0.5 (ρ = %s)", rho_low),
    sprintf("Conditional: Z > 0.5 (ρ = %s)", rho_high),
    sprintf("Partial copula (ρ = %s)", rho_all)
  )))

ggplot(plot_data_11c, aes(x = U_X, y = U_Y)) +
  geom_point(aes(color = Z), alpha = 0.4, size = 0.6) +
  scale_color_viridis_c(option = "viridis", name = "Z") +
  facet_wrap(~ Panel, ncol = 3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = expression(U[X]), y = expression(U[Y])) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
```

# Discussion

In this paper, we discussed the notion of partial copula, originally introduced in @bergsma2004 to test conditional independence. This test was introduced upon the observation that, under conditional independence, the variables $F_{X|Z}(X|Z)$ and $F_{Y|Z}(Y|Z)$, usually known as Rosenblatt transformations, are independent. Although we recognize the usefulness of the partial copula as a tool to test conditional independence, our main focus has been to highlight its broader applicability as a general representation of conditional association, which has not been well explored previously in the literature, to the best of our knowledge. The starting point from this paper is the observation that the variables $F_{X|Z}(X|Z)$ and $F_{Y|Z}(Y|Z)$ are uniformly distributed in $[0,1]$, which means that their joint cumulative distribution function is a copula, which enables studying the several dependence concepts constructed in the copula literature, such as Spearman's and Kendall's coefficients.

First, we showed in Proposition 3 and Remark 2 that the partial copula of $(X,Y)$ adjusted by $Z$ may be viewed as the joint distribution of $(X,Y)$ under the assumption that $X$ and $Y$ are both independent of $Z$ (marginally) but preserving their conditional copula. This interpretation clarifies the role of the partial copula as a general representation of conditional association, since it captures the dependence between $X$ and $Y$ after removing the effect of $Z$. Second, we established several properties of the partial copula, including bounds for Spearman's and Kendall's coefficients in the partial copula when the maximum value of the Kolmogorov distance dependence in the conditional copulas is known (Propositions 6 and 7), the connection between conditional and partial correlations (Proposition 8), and a characterization of the partial copula when the conditional copulas belongs to a specific parametric family (Proposition 9). In an intuitive level, those results show that the partial copula captures the properties of conditional copulas while removing the effect of $Z$.

In the context of causal inference, if $Z$ is a confounding variable, the dependence between $X$ and $Y$ adjusted by $Z$ - as traditionally estimated in linear models and other statistical techniques - may be interpreted as the true causal connection between the two variables. The properties proved in this paper underscore the capability of the partial copula to capture this causal connection without requiring restrictive assumptions such as linearity or the simplifying assumption used in vine copula modeling.

Finally, a simulation study was conducted to compare how the partial and the marginal copulas may be different. In scenarios presenting conditional independence (scenarios 2, 3 and 4), the partial copula is the independence copula, while the marginal copula contains the spurious association induced by $Z$. In scenarios where the confounding effect reinforces the conditional dependence (1, 6, and 9), the association observed in the marginal copula was stronger than in the partial copula. In scenarios where the confounding effect opposes the conditional dependence (5, 7, and 8), Simpson's paradox arises: marginal and partial correlations have opposite signs. In scenarios with no confounding (10), the marginal and the partial copula are the same. Finally, scenario 11 illustrated cases violating the simplifying assumption, which shows a fundamental difference between partial and conditional copulas: conditional copulas depend on the specification of the value being conditioned in the variable $Z$, while the partial copula does not. In this sense, the partial copula may be seen as a "summary" of the conditional copulas across all values of $Z$.

The diversity of scenarios simulated shows that, in several contexts, the marginal association and the average conditional association may be very different - including scenarios known as Simpson's paradox, where even the signs of correlation coefficients are different. In causal inference literature [@pearl2009; @pearl2016], it is well known that marginal associations may be very different from causal effects due to confounding. The results in this paper show that the partial copula is a useful tool to estimate the average conditional association, which may be interpreted as a causal effect under standard assumptions in causal inference. From this paper, we hope to have showed both the importance of using conditional association instead of marginal association in several contexts, and the capability of the partial copula to capture this conditional association in a general way.

Some directions for future research include studying directed dependence coefficients [@azadkia2021simple; @fuchs] in the context of partial copulas, and exploring the use of partial copulas in the context of causal discovery algorithms. Furthermore, further advances on the estimation of partial copulas from real data is of great interest for the scientific community, since it would allow to apply the insights of this paper in empirical contexts.

\newpage

\bibliography{references}

\newpage

# Appendix {-}

```{r scatter-fn, echo=FALSE}
create_scenario_scatter <- function(data, scenario_num) {

  scen_data <- data %>% filter(Scenario == scenario_num)

  case_order <- scen_data %>%
    distinct(Case, Case_Label) %>%
    arrange(Case) %>%
    pull(Case_Label)

  plot_data <- scen_data %>%
    select(Case_Label, X, Y, U_X, U_Y, Z, rho_XY, rho_UV, tau_XY, tau_UV) %>%
    pivot_longer(cols = c(X, U_X), names_to = "x_type", values_to = "x_val") %>%
    pivot_longer(cols = c(Y, U_Y), names_to = "y_type", values_to = "y_val") %>%
    filter((x_type == "X" & y_type == "Y") | (x_type == "U_X" & y_type == "U_Y")) %>%
    mutate(
      Type = ifelse(x_type == "X", "Marginal", "Partial"),
      Type = factor(Type, levels = c("Marginal", "Partial")),
      Case_Label = factor(Case_Label, levels = case_order)
    )

  annot_data <- scen_data %>%
    distinct(Case_Label, rho_XY, rho_UV, tau_XY, tau_UV) %>%
    pivot_longer(
      cols = c(rho_XY, rho_UV, tau_XY, tau_UV),
      names_to = "measure",
      values_to = "value"
    ) %>%
    mutate(
      Type = ifelse(grepl("UV", measure), "Partial", "Marginal"),
      Type = factor(Type, levels = c("Marginal", "Partial")),
      Coef = ifelse(grepl("rho", measure), "rho", "tau"),
      Case_Label = factor(Case_Label, levels = case_order)
    ) %>%
    select(-measure) %>%
    pivot_wider(names_from = Coef, values_from = value) %>%
    mutate(label = sprintf("\u03C1=%.2f, \u03C4=%.2f", rho, tau))

  ggplot(plot_data, aes(x = x_val, y = y_val)) +
    geom_point(aes(color = Z), alpha = 0.4, size = 0.6) +
    geom_text(data = annot_data, aes(x = 0.95, y = 0.05, label = label),
              hjust = 1, vjust = 0, size = 2.5, color = "black") +
    scale_color_viridis_c(option = "viridis", name = "Z") +
    facet_grid(Case_Label ~ Type, scales = "fixed") +
    labs(x = NULL, y = NULL) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    theme_minimal(base_size = 9) +
    theme(
      strip.text = element_text(face = "bold", size = 8),
      strip.text.y = element_text(angle = 0, hjust = 0),
      legend.position = "right",
      panel.grid.minor = element_blank()
    )
}
```

```{r scatter-1, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 1 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$. Each panel displays $\\rho$ (Spearman) and $\\tau$ (Kendall). Color gradient represents $Z$."}
create_scenario_scatter(all_data, 1)
```

```{r scatter-2, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 2 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$, where ind denotes independence."}
create_scenario_scatter(all_data, 2)
```

```{r scatter-3, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 3 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$, where ind denotes independence. (R) indicates 90-degree rotation."}
create_scenario_scatter(all_data, 3)
```

```{r scatter-4, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 4 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$, where ind denotes independence."}
create_scenario_scatter(all_data, 4)
```

```{r scatter-5, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 5 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$."}
create_scenario_scatter(all_data, 5)
```

```{r scatter-6, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 6 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$. (R) indicates 90-degree rotation."}
create_scenario_scatter(all_data, 6)
```

```{r scatter-7, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 7 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$. (R) indicates 90-degree rotation."}
create_scenario_scatter(all_data, 7)
```

```{r scatter-8, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 8 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$."}
create_scenario_scatter(all_data, 8)
```

```{r scatter-9, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 9 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$."}
create_scenario_scatter(all_data, 9)
```

```{r scatter-10, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 10 (see Table 1). Each row corresponds to a copula family and parameter triplet $(\\theta_{XZ}, \\theta_{YZ}, \\theta_{XY|Z})$, where ind denotes independence."}
create_scenario_scatter(all_data, 10)
```

```{r scatter-11, echo=FALSE, dev="png", dpi=300, out.width="100%", fig.cap="Scenario 11 (see Table 1). Simplifying assumption violated. The conditional copula parameter $\\theta(z)$ varies with $Z$."}
create_scenario_scatter(all_data, 11)
```
